---
title: "chad_huntebrinker_hw3"
author: "Chad Huntebrinker"
date: "2024-09-14"
output:
  word_document: default
  html_document: default
---

Part 1: Basic Analysis of Cancer Patients Data Using Lists
  
Question 1: create a list object
Create a list named patient_data to represent information about cancer patients.
Include the given 5 patients, within the following four vectors on the list:
  
PatientID (Numeric) : 1, 2, 3, 4, 5
Age (Numeric) : 45, 32, 58, 41, 50
CancerType (Character) : "Breast", "Lung", "Colon", "Breast", "Lung"
TreatmentStatus (Logical) : TRUE, FALSE, TRUE, TRUE, FALSE

Display the patient_data.
```{r}
patient_data <- list(
  PatientID = c(1, 2, 3, 4, 5),
  Age = c(45, 32, 58, 41, 50),
  CancerType = c("Breast", "Lung", "Colon", "Breast", "Lung"),
  TreatmentStatus = c(TRUE, FALSE, TRUE, TRUE, FALSE))
patient_data
```

Question 2 : Perform Basic Operations on the list object
Find the average age of patients.
```{r}
attributes(patient_data)
length(patient_data)
class(patient_data)
print(mean(patient_data$Age))
```

Retrieve the patientIDs with “Breast” cancer type.
```{r}
index <- which(patient_data$CancerType == "Breast")
ids <- patient_data$PatientID[index]
print(ids)
```


Update the treatment status for the “Colon” cancer patient to FALSE.
```{r}
patient_data$TreatmentStatus[3] <- FALSE
patient_data$TreatmentStatus[patient_data$CancerType=="Colon"]

```

Display the updated patient_data.
```{r}
print(patient_data)
```

Calculate the percentage of patients undergoing treatment
```{r}
num_treatments <- which(patient_data$TreatmentStatus == TRUE)
final_percentage <- (length(num_treatments) / 5) * 100
cat(toString(final_percentage), "%", sep = "")
```

Question 3: Update the patient_data by adding another patient with the following details.
Print the updated list

```{r}
patient_data$PatientID[6] <- 6 
patient_data$Age[6] <- 68
patient_data$CancerType[6] <- "Brain"
patient_data$TreatmentStatus[6] <- TRUE
print(patient_data)
```

Question 4: plot a histogram for the age of patients
```{r}
hist(patient_data$Age, xlab = "Age", main = "Histogram of Patient Ages")
```

------------------------------------------End of Part 1 --------------------------------------------------------
  
Part 2: Analysis of Breast cancer data set
You are provided with a breast cancer data set (BC_data.cvs) and you have to work on this table as per the instructions provided.

Question 1:
Read the Breast cancer data into a dataframe called "Bc.data"
```{r}
rm(list = ls())
Bc.data <- read.csv("BC_data.csv", header = TRUE, row.names = 1)
```

The rownames should be the patient_IDs.
```{r}
rownames(Bc.data)
```

Perform all the basic functions on the dataframe and get familiarize with the data.
```{r}
dim(Bc.data)
class(Bc.data)
str(Bc.data)
ncol(Bc.data)
colnames(Bc.data)
```

Question 2:
Check if NA values are available.
```{r}
any(is.na(Bc.data))
```

create a vector called "Surv_Mnths" with the data from "Survival_Months" column
```{r}
Surv_Mnths <- Bc.data$Survival_Months
```

create a vector called "Tum_Size" with the data from "Tumor_Size" column
```{r}
Tum_Size <- Bc.data$Tumor_Size
```

check the correlation between "Surv_Mnths" and "Tum_Size" and interpret the relation based on the correlation value. 
Write your interpretation in the comments. Clue: There is a built-in function for correlation
```{r}
cor(Surv_Mnths, Tum_Size)
#The cor function returns a value of -0.1989186
#Thus, there is a negative linear relationship between the two
```

Calculate the mean and median of the "Survival_Months" column.
```{r}
mean(Surv_Mnths)
median(Surv_Mnths)
```

generate a scatter plot "Surv_Mnths" Vs "Tum_Size" and present with proper axis labels
```{r}
plot(Surv_Mnths, Tum_Size, ylab= "Tumor Size", xlab = "Survival Months", main = "Tumor Size to Survival Months")
```

Question 3:
Print a table to show how many patients are DECEASED and how many LIVING
```{r}
table(Bc.data$Survival_Status)
```

Create a new dataframe called "living_patient" with only living patients.
```{r}
living_patients <- subset(Bc.data, Bc.data$Survival_Status=="LIVING")
```
In these living patients, how many are undergoing Chemotherapy and how many undergoing Hormone_Therapy
```{r}
table(living_patients$Chemotherapy)
table(living_patients$Hormone_Therapy)
```


Question 4:
  
Create an additional column in the BC.data dataframe with the name 'Full_Info' by
concatenating the values of Survival_Status with Tumor_Stage seperated by underscore.
```{r}
Bc.data$Full_Info <- paste(Bc.data$Survival_Status, Bc.data$Tumor_Stage, sep = "_")
```

Create additional column 'Survival_Status_Short' by replacing the  the Survival_Status values "DECEASED" with "D" and "LIVING" with "L".
```{r}
Bc.data$Survival_Status_Short[Bc.data$Survival_Status == "LIVING"] <- "L"
Bc.data$Survival_Status_Short[Bc.data$Survival_Status == "DECEASED"] <- "D"
```

Check the factor levels of Survival_Status, Tumor_Stage  and Surgery_Type in BC.data dataframe
```{r}
factor(Bc.data$Survival_Status)
factor(Bc.data$Tumor_Stage)
factor(Bc.data$Surgery_Type)
```
