---
title: "Homework 7"
author: "Chad Huntebrinker"
date: "2024-10-10"
output: word_document
---
Question 1
A function is defined in the following chunk to calculate the square of each element using a for loop. Convert this function into a vectorized function
```{r}
# Define a function to calculate the square of each element using a for loop
square_with_loop <- function(x) {
  result <- numeric(length(x))  # Initialize an empty numeric vector to store results
  for (i in seq_along(x)) {
    result[i] <- x[i]^2  # Calculate square of each element and store in result vector
  }
  return(result)
}

# Test the function
numbers <- c(1, 2, 3, 4, 5)
print(square_with_loop(numbers))

```

Answer to Question 1
```{r}
square_with_loop <- function(x) {
  return(x^2)
}

# Test the function
numbers <- c(1, 2, 3, 4, 5)
print(square_with_loop(numbers))

```

Question 2
A function is defined in the following chunk  to calculate the cumulative sum of elements in a given numeric vector using a for loop.Convert this function into a vectorized function
```{r}
# Define function to calculate cumulative sum using a for loop
cumulative_sum_for_loop <- function(x) {
  result <- numeric(length(x))  # Initialize a numeric vector to store cumulative sum
  sum_so_far <- 0  # Initialize variable to keep track of sum so far
  for (i in seq_along(x)) {
    sum_so_far <- sum_so_far + x[i]  # Update sum so far
    result[i] <- sum_so_far  # Store cumulative sum in result vector
  }
  return(result)
}

# Test the function
numbers <- c(1, 2, 3, 4, 5)
print(cumulative_sum_for_loop(numbers))

```

Answer to Question 2
```{r}
# Define function to calculate cumulative sum using a for loop
cumulative_sum_for_loop <- function(x) {
  return(sum(x))
}

# Test the function
numbers <- c(1, 2, 3, 4, 5)
print(cumulative_sum_for_loop(numbers))

```

Question 3
Perform k-means clustering on the Iris dataset both serially and in parallel. Perform clustering on the first four columns of the Iris dataset, excluding the species (5th) column.
report the computation times
```{r}
library(datasets)
library(parallel)
data(iris)
fx <- function(nstart){
  # Perform k-means clustering on a data matrix Boston.
  return(kmeans(iris_data, centers = 4, nstart = nstart))
}
#Get the first four columns
iris_data <- data.frame(iris$Sepal.Length, iris$Sepal.Width, iris$Petal.Length, iris$Petal.Width)
starts <- rep(100, 400)

#Complete serieally
system.time(results <- lapply(starts, fx))

#Complete parallel
numCores <- detectCores()
cl <- makeCluster(8)
clusterExport(cl=cl, c('iris_data'))
results <- parLapply(cl,starts, fx)
system.time(results <- parLapply(cl,starts, fx))
stopCluster(cl)

```
