---
title: "chad_huntebrinker_hw1"
author: "Chad Huntebrinker"
date: "2024-08-31"
output: word_document
---

Question 1: Calculate the mean, median, and standard deviation of the following vector.
my_vec <- c(15, 20, 25, 30, 35)

```{r}
my_vec <- c(15, 20, 25, 30, 35)
mean_vec <- mean(my_vec)
median_vec <- median(my_vec)
standard_vec <- sd(my_vec)

#Print vector and it's mean, median, and standard deviation
print(my_vec)
print(mean_vec)
print(median_vec)
print(standard_vec)
```

Question 2: Generate two vectors called “abc” and “xyz” with a random values of 100 in each. Generate scatter plot, bar plot and line plot with required labels and headings.

```{r}
#Generate two vectors with random values between 1 and 20
abc <- runif(100, min = 1, max = 20)
xyz <- runif(100, min = 1, max = 20)

#Plot the scatter, bar, and line plot
plot(abc, xyz, xlab = "abc axis", ylab = "xyz axis", main = "Scatter Plot")
barplot(abc, xyz, xlab = "abc axis", ylab = "xyz axis", main = "Bar Plot")
plot(abc, xyz, type = "l", xlab = "abc axis", ylab = "xyz axis", main = "Line Plot")
```

Question 3: Generate 1000 random numbers from a normal distribution with mean 0 and standard deviation 10. Save the result into object x. Plot a histogram for x object. Calculate the mean and variance of x.

```{r}
#create vector and histogram
x <- rnorm(1000, mean = 0, sd = 10)
hist(x)

#print mean and variance of vector x
print(mean(x))
print(var(x))
```

