---
title: "Linear Regression"
output: word_document
---

```{r setup}
library(tidyverse)
library(modelr)
library(broom)
library(readxl)

# Import wages here
wages <- read_excel("wages-1.xlsx", na = "NA")

# Fall back in case you cannot load wages
# wages <- heights %>%
#   filter(income > 0) %>%
#   mutate(marital = as.character(marital),
#          sex = as.character(sex))
```

## Your Turn 1

* Save the `wages.xlsx` dataset to your computer.  
* Change the working directory to the same location.
* Import `wages.xlsx` as wages and *copy the code to your setup chunk*.
* Be sure to set NA: to NA.

## Your Turn 2

Fit the model on the slide and then examine the output. What does it look like?

```{r}
mod_e <- lm(log(income)~ education, data = wages)
mod_e
class(mod_e)
```

## Your Turn 3

Use a pipe to model `log(income)` against `height`. Then use broom and dplyr functions to extract:

1. The **coefficient estimates** and their related statistics 
2. The **adj.r.squared** and **p.value** for the overall model

```{r}
mod_h <- wages %>% 
  lm(log(income)~height, data =.)

mod_h %>% tidy()

mod_h %>% 
  glance() %>% 
  select(adj.r.squared, p.value)
```


## Your Turn 4

Model `log(income)` against `education` and `height` and `sex`. Interpret the coefficients in narrative here.

```{r}
mod_ehs <- wages %>% 
  lm(log(income)~education + height + sex, data = .)

mod_ehs %>% tidy()
#The sexmale term is an indicator variable (0,1).  As a result, if the person is a male they get 0.46 but
#if they are a female, they get 0.  We also see that taller people are expected to get paid 
#slightly more than shorter ones.

```

## Your Turn 5

Add `+ geom_smooth(method = lm)` to the code below. What happens?

```{r}
wages %>%
     ggplot(aes(x = height, y = log(income))) +
     geom_point(alpha = 0.1) + 
     geom_smooth(method=lm)
#geom_smooth adds a line for predicting how height impacts income.
```


## Your Turn 6

Use `add_predictions` to make the plot below.

```{r warning = FALSE, message = FALSE}

# In case you haven't made the ehs model
mod_ehs <- wages %>% lm(log(income) ~ education + height + sex, data = .)

# Make plot here
wages %>% 
  add_predictions(mod_ehs) %>% 
  ggplot(mapping = aes(x = height, y = pred, color = sex)) +
  geom_line() + 
  facet_wrap(~ education)
```

***

# Take Aways

* Use `glance()`, `tidy()`, and `augment()` from the **broom** package to return model values in a data frame.

* Use `add_predictions()` or `gather_predictions()` or `spread_predictions()` from the **modelr** package to visualize predictions.

* Use `add_residuals()` or `gather_residuals()` or `spread_residuals()` from the **modelr** package to visualize residuals.
